---
title: "Introduction to CARVE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CARVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

CARVE chooses the number of clusters, and the clustering method, by
resampling. Each candidate configuration (a clustering method, its
parameters and a number of clusters) clusters many subsamples of the data
and gets two scores. Stability is the agreement between the clusterings of
two overlapping subsamples on the samples they share. Generalizability is
how well a random forest trained on one subsample's clusters predicts the
clusters of the samples held out of it. Both are adjusted Rand indices
(ARI), averaged over the resamples. A selection rule picks a configuration
from the scores, and the labels come from the consensus of that
configuration's resamples.

This vignette fits CARVE to simulated data and goes through the results.
`vignette("single-cell", package = "CARVE")` covers SingleCellExperiment
and Seurat objects, `vignette("customizing", package = "CARVE")` the
options of a run, and `vignette("python-users", package = "CARVE")` the
differences from the Python package.

## Simulated data

Four groups of 50 points in two dimensions:

```{r data, fig.width = 4.5, fig.height = 4}
library(CARVE)

set.seed(1)
centers <- rbind(c(0, 0), c(4, 0), c(2, 3.5), c(6, 3.5))
truth <- rep(1:4, each = 50)
X <- centers[truth, ] + matrix(rnorm(400, sd = 0.8), ncol = 2)
plot(X, col = truth, pch = 19, asp = 1, xlab = "", ylab = "")
```

## Fitting

```{r fit}
fit <- carve(X, n_clusters = 2:8, n_resamples = 30, random_state = 1)
fit
```

`carve()` takes a numeric matrix or data frame with one row per sample.
`n_clusters` lists the numbers of clusters to compare. By default CARVE
compares three methods, k-means, Ward's agglomerative clustering and
spectral clustering, so this run has
`r nrow(estimator_results(fit))` configurations. Each configuration is
clustered on `n_resamples` resamples. The default is 100; fewer resamples
give noisier scores. Resample `b` draws its subsamples with the seeds
`random_state + b` and `random_state + b + n_resamples`, where the default
`random_state = NULL` counts as 0. Each resample's seeds depend only on its
index, so a run gives the same results on any number of cores.

## The results table

```{r results}
results <- estimator_results(fit)
head(results[, c("config_id", "method_label", "n_clusters", "ari_stability",
                 "ari_stability_se", "ari_generalizability")])
```

Each row is one configuration, identified by its `config_id`.
`ari_stability` and `ari_generalizability` are means over the resamples,
with their standard errors in the `_se` columns and their 5th and 95th
percentiles in the `_lower` and `_upper` columns. `ari_average` is the
mean of the two. The table also has three scores computed from each
configuration's consensus matrix (`consensus_pac_stability`,
`consensus_gini_stability` and `consensus_ce_stability`) and the
classifier's accuracy on the held-out samples
(`accuracy_generalizability`).

## Selecting a configuration

```{r select}
get_k(fit)
get_k(fit, measure = "generalizability", rule = "max")
get_k(fit, measure = "gini", rule = "max")
```

`measure` names the score: `"stability"` (the default),
`"generalizability"`, `"average"`, or one of the consensus and accuracy
scores, `"pac"`, `"gini"`, `"ce"` and `"accuracy"`. `rule` says how to
pick a configuration:

- `"max"` takes the one with the highest mean score.
- `"1se"`, the default, takes the largest number of clusters among the
  configurations whose score is within one standard error of the highest.
- `"quantile"` does the same with another band: the 5th to 95th
  percentiles of the best configuration's resample scores.

The consensus and accuracy scores have no standard errors or percentiles.
With them, `"1se"` and `"quantile"` fall back to `"max"` with a warning, so
pass `rule = "max"`.

`get_sweep_value()` returns the selected value of the swept parameter,
here the number of clusters again. `get_estimator()` returns the selected
method as a function of the data and a seed, with its parameters set:

```{r estimator}
get_sweep_value(fit)
estimator <- get_estimator(fit)
attr(estimator, "estimator")
attr(estimator, "params")
table(estimator(X, random_state = 1))
```

## Two clusters

Stability can be high at two clusters even when the data have more
structure, because a split into two halves is easy to repeat.
`not_two = TRUE` leaves two clusters out of the selection. Here two
clusters score lowest, so the choice does not change:

```{r not-two}
get_k(fit, not_two = TRUE)
```

## Labels

```{r labels}
labels <- get_labels(fit)
table(labels, truth)
```

`get_labels()` does not fit the method again to all the data. It takes
the selected configuration's consensus matrix, whose entry for two samples
is the share of the subsamples drawing both in which they fell in the same
cluster, and cuts an average-linkage tree built from it. The labels run
from 1 to the number of clusters. `k` asks for the best configuration at
another number of clusters:

```{r labels-k}
table(get_labels(fit, k = 3), truth)
```

## Plots

Every plot is a ggplot object, so `+` changes it and `ggplot2::ggsave()`
saves it.

```{r metric, fig.width = 7}
plot_metric_over_n_clusters(fit)
plot_metric_over_n_clusters(fit, measure = "generalizability") +
  ggplot2::labs(title = "Four simulated groups")
```

Each line is one method, with standard error bars, and the dashed line
marks the selected configuration.

```{r consensus, fig.width = 5, fig.height = 4.5}
plot_consensus_matrix(fit)
```

The heatmap is the selected configuration's consensus matrix, with the
samples ordered by cluster and the clusters shown in the band above it. A
cluster that the resamples reproduce is a block of values near 1 on the
diagonal.

Four plots show per-sample scores, and `source` picks the score.
`"gini"`, the default, and `"ce"` measure stability. Both run from 0 to 1,
and a sample scores 1 when every other sample drawn with it fell in its
cluster either every time or never. `"accuracy"` measures generalizability: the share of the
draws holding the sample out in which the classifier predicted its
cluster.

```{r per-sample}
plot_cluster_violin(fit)
plot_cluster_boxplot(fit, source = "accuracy")
```

The two scatter plots draw the samples over the data the fit was given,
or over the first two columns of `embedding`, such as a UMAP. Samples with
low scores are drawn larger and more opaque. In the diagnostic scatter the
shape gives the cluster and the fill the score.

```{r scatter, fig.width = 6, fig.height = 4.5}
plot_cluster_scatter(fit)
plot_diagnostic_scatter(fit, source = "ce")
```

## Scores and matrices of one configuration

`config_id` looks up a configuration's per-sample scores and consensus
matrix. Here it is the selected configuration's:

```{r scores}
id <- results$config_id[results$estimator == attr(estimator, "estimator") &
                          results$n_clusters == get_k(fit)]
id
summary(sample_scores(fit, id, source = "gini"))
dim(consensus_matrix(fit, id))
```

The fit also keeps the sweep, the grids it ran and the data:

```{r accessors}
sweep_spec(fit)
estimator_param_grids(fit)[[1]]
dim(input_data(fit))
```

## Saving a fit

A fit is an ordinary R object, saved with `saveRDS()` and read back with
`readRDS()`. It holds two consensus matrices of `n` by `n` doubles for
each configuration, so the file grows with the square of the number of
samples.

```{r save}
path <- tempfile(fileext = ".rds")
saveRDS(fit, path)
identical(get_labels(readRDS(path)), labels)
```

## Session information

```{r session}
sessionInfo()
```
