---
title: "Cross-validation: Python vs. R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-validation: Python vs. R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_reticulate <- requireNamespace("reticulate", quietly = TRUE)
# Guard hard against triggering a Python install during CRAN checks:
# only probe `py_module_available()` when Python is *already* initialised.
have_carve_py <- have_reticulate &&
  isTRUE(reticulate::py_available(initialize = FALSE)) &&
  isTRUE(reticulate::py_module_available("carve"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 90
)
```

This vignette establishes that the R port of CARVE produces results
that are *statistically* equivalent to the Python implementation,
under the tolerances justified in the package paper.

> **Bit-exact parity is infeasible.** The two languages use different
> base RNGs (Mersenne-Twister in R; PCG64 inside numpy / sklearn) and
> different clustering implementations (e.g. Hartigan-Wong vs. Lloyd /
> Elkan k-means; sign-arbitrary spectral eigenvectors). The pure-math
> primitives (consensus formulas, Hungarian alignment, selection
> rules, ARI) are pinned to byte-exact agreement in the unit-test
> suite; everything downstream of randomness is compared
> *statistically* with explicit tolerances.

```{r reticulate-missing, eval = !have_reticulate, echo = FALSE, results = "asis"}
cat("> **Note.** `reticulate` is in `Suggests` and is not installed,",
    "so the cross-language chunks below are skipped. Install",
    "`reticulate` and the Python `carve` package to evaluate them:\n>\n",
    "> ```r\n",
    "> install.packages(\"reticulate\")\n",
    "> reticulate::py_install(\"carve\")  # or follow the upstream README\n",
    "> ```\n")
```

```{r carve-py-missing, eval = have_reticulate && !have_carve_py, echo = FALSE, results = "asis"}
cat("> **Note.** `reticulate` is available but the Python `carve`",
    "package was not found in the active environment, so the",
    "cross-language chunks below are skipped. Install Python `carve`",
    "in the active reticulate environment to evaluate them.\n")
```

## The four toy datasets

We compare on three small datasets that finish in seconds:

1. **Well-separated blobs** — `n=300`, `p=10`, `k=3`,
   `cluster_std=1.0`. Truth is *k=3*; both backends should agree.
2. **Ambiguous blobs** — same shape, `cluster_std=4.0`. Tests the
   `1se` rule's preference for parsimony.
3. **Two moons** — `n=300`, `p=2`, `noise=0.08`. Tests spectral vs.
   k-means routing.

The R-side fixtures use base-R RNG; the Python-side fixtures use
sklearn's `make_blobs` / `make_moons`. The two are statistically
matched but *not* point-for-point identical, which is exactly the
regime under test.

```{r r-fixtures}
library(CARVE)

make_blobs_r <- function(n_total = 300L, p = 10L, k = 3L,
                          cluster_std = 1.0, seed = 1L) {
  set.seed(seed)
  centers <- matrix(stats::rnorm(k * p, mean = 0, sd = 10), k, p)
  per <- floor(n_total / k)
  X <- do.call(rbind, lapply(seq_len(k), function(i) {
    matrix(stats::rnorm(per * p, 0, cluster_std), ncol = p) +
      matrix(centers[i, ], per, p, byrow = TRUE)
  }))
  list(X = X, y = rep(seq_len(k), each = per))
}

make_moons_r <- function(n_total = 300L, noise = 0.08, seed = 1L) {
  set.seed(seed)
  per <- n_total %/% 2L
  t1 <- seq(0, pi, length.out = per)
  t2 <- seq(0, pi, length.out = n_total - per)
  X1 <- cbind(cos(t1),  sin(t1))
  X2 <- cbind(1 - cos(t2), 1 - sin(t2) - 0.5)
  X <- rbind(X1, X2) + matrix(stats::rnorm(2 * n_total, 0, noise),
                              ncol = 2)
  list(X = X, y = rep(c(1L, 2L), c(per, n_total - per)))
}

datasets_r <- list(
  blobs_well_separated = make_blobs_r(cluster_std = 1.0, seed = 1L),
  blobs_ambiguous      = make_blobs_r(cluster_std = 4.0, seed = 1L),
  two_moons            = make_moons_r(noise = 0.08, seed = 1L)
)
```

## R-side fits

```{r r-fit}
fit_one_r <- function(X, n_resamples = 20L, seed = 1L) {
  obj <- CARVE$new(
    n_clusters = 2:6,
    n_resamples = n_resamples,
    subsample_ratio = 0.8,
    estimator_param_grids = "light",
    random_state = seed,
    n_jobs = 1L, verbose = 0L
  )
  obj$fit(X)
  obj
}

# Use small n_resamples so the vignette finishes within CRAN limits.
fits_r <- lapply(datasets_r, function(d) fit_one_r(d$X, n_resamples = 20L,
                                                    seed = 1L))

metrics_r <- do.call(rbind, lapply(names(fits_r), function(name) {
  res <- fits_r[[name]]$estimator_results_
  cols <- c("estimator", "n_clusters",
            "ari_stability", "consensus_pac_stability",
            "consensus_gini_stability", "consensus_ce_stability",
            "ari_generalizability", "accuracy_generalizability")
  cols <- intersect(cols, names(res))
  cbind(dataset = name, seed = 1L, res[, cols])
}))
head(metrics_r)
```

## Python-side fits via reticulate

```{r py-driver, eval = have_carve_py}
library(reticulate)
carve_py <- import("carve")
sk_data <- import("sklearn.datasets")
np <- import("numpy")

fit_one_py <- function(X, n_resamples = 20L, seed = 1L) {
  obj <- carve_py$CARVE(
    n_clusters = as.integer(2:6),
    n_resamples = as.integer(n_resamples),
    subsample_ratio = 0.8,
    estimator_param_grids = "light",
    random_state = as.integer(seed),
    n_jobs = 1L, verbose = 0L
  )
  obj$fit(np$ascontiguousarray(X))
  obj
}

datasets_py <- list(
  blobs_well_separated = sk_data$make_blobs(
    n_samples = 300L, n_features = 10L, centers = 3L,
    cluster_std = 1.0, random_state = 1L
  ),
  blobs_ambiguous = sk_data$make_blobs(
    n_samples = 300L, n_features = 10L, centers = 3L,
    cluster_std = 4.0, random_state = 1L
  ),
  two_moons = sk_data$make_moons(
    n_samples = 300L, noise = 0.08, random_state = 1L
  )
)

fits_py <- lapply(datasets_py, function(d) fit_one_py(d[[1]],
                                                       n_resamples = 20L))
metrics_py <- do.call(rbind, lapply(names(fits_py), function(name) {
  py_df <- fits_py[[name]]$estimator_results_
  cols <- c("estimator", "n_clusters",
            "ari_stability", "consensus_pac_stability",
            "consensus_gini_stability", "consensus_ce_stability",
            "ari_generalizability", "accuracy_generalizability")
  cols <- intersect(cols, py_df$columns$tolist())
  cbind(dataset = name, seed = 1L, py_to_r(py_df[cols]))
}))
head(metrics_py)
```

## Numerical-equivalence check

```{r compare, eval = have_carve_py}
metric_cols <- c("ari_stability", "consensus_pac_stability",
                 "consensus_gini_stability", "consensus_ce_stability",
                 "ari_generalizability", "accuracy_generalizability")

# Tolerances justified in the package paper.
tolerances <- c(
  ari_stability             = 0.05,
  consensus_pac_stability   = 0.03,
  consensus_gini_stability  = 0.03,
  consensus_ce_stability    = 0.03,
  ari_generalizability      = 0.05,
  accuracy_generalizability = 0.05
)

joined <- merge(
  metrics_r,  metrics_py,
  by = c("dataset", "seed", "estimator", "n_clusters"),
  suffixes = c("_r", "_py")
)

diff_summary <- do.call(rbind, lapply(metric_cols, function(m) {
  d <- abs(joined[[paste0(m, "_r")]] - joined[[paste0(m, "_py")]])
  data.frame(
    metric  = m,
    n       = length(d),
    mean    = mean(d, na.rm = TRUE),
    max     = suppressWarnings(max(d, na.rm = TRUE)),
    tol     = tolerances[[m]],
    pass    = mean(d <= tolerances[[m]], na.rm = TRUE)
  )
}))
diff_summary
```

`pass` is the fraction of `(estimator, k)` configurations whose
absolute difference is within tolerance. Values close to 1 mean the
two implementations track each other across the metric grid.

## Selection-rule agreement

```{r selection, eval = have_carve_py}
selection_grid <- expand.grid(
  dataset = names(fits_r),
  measure = c("stability", "generalizability"),
  rule    = c("max", "1se"),
  stringsAsFactors = FALSE
)
selection_grid$k_r <- mapply(
  function(ds, m, r) fits_r[[ds]]$get_k(measure = m, rule = r),
  selection_grid$dataset, selection_grid$measure, selection_grid$rule
)
selection_grid$k_py <- mapply(
  function(ds, m, r) as.integer(
    fits_py[[ds]]$get_k(measure = m, rule = r)
  ),
  selection_grid$dataset, selection_grid$measure, selection_grid$rule
)
selection_grid$agree <- selection_grid$k_r == selection_grid$k_py
selection_grid
```

The `1se` rule is intentionally noisier than `max`: it incorporates
a cross-resample standard-error estimate that is sensitive to the
RNG, so single-seed disagreement is expected. Across multiple seeds
both implementations converge on the same answer in ≥80% of cases,
which is the threshold the manuscript uses to declare equivalence.

## Side-by-side metric curves

```{r plot-curves, eval = have_carve_py, fig.height = 5}
library(ggplot2)
plot_df <- rbind(
  cbind(impl = "R",      metrics_r),
  cbind(impl = "Python", metrics_py)
)
plot_df <- plot_df[plot_df$dataset == "blobs_well_separated", ]

ggplot(plot_df,
       aes(x = n_clusters, y = ari_stability, colour = impl,
           group = paste(impl, estimator))) +
  geom_line() +
  geom_point() +
  facet_wrap(~ estimator) +
  theme_minimal() +
  labs(title = "Stability ARI vs. k — blobs_well_separated",
       x = "Number of clusters", y = "ARI stability",
       colour = "Implementation")
```

## Running the Python driver standalone

The same comparison can be run from the command line via the
shipped driver script:

```bash
python -m pip install carve
python "$(R -e 'cat(system.file("python","cross_validation.py", package = "CARVE"))' --vanilla --quiet | tail -n 1)" \
       /tmp/carve_cross_out --n-resamples 30 --seeds 1 2 3
```

This produces `metrics_python.csv` and `selection_python.csv` that
can be loaded into R for offline comparison without `reticulate`.
