---
title: "Customizing a CARVE run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing a CARVE run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

This vignette goes through the arguments of `carve()` that change what a
run compares and how it runs: the methods and their parameters, a
clustering function of your own, density-based clustering with noise, the
classifier, randomized preprocessing, reference labels, and parallel runs.
`vignette("CARVE", package = "CARVE")` introduces the scores and the
selection rules.

The data are four groups of 60 points in ten dimensions:

```{r data}
library(CARVE)

set.seed(1)
truth <- rep(1:4, each = 60)
centers <- matrix(rnorm(40, sd = 2), nrow = 4)
X <- centers[truth, ] + matrix(rnorm(240 * 10), ncol = 10)
```

## Methods and parameters

The clustering methods are plain R functions that take the data and their
parameters and return one label per row:

```{r kmeans}
table(KMeans(X, n_clusters = 4, random_state = 0), truth)
```

`estimator_grid()` pairs a function with the values to try for each of its
arguments, and `carve()` runs every combination:

```{r grids}
grids <- list(
  estimator_grid(KMeans, n_clusters = 2:6, n_init = c(1, 10)),
  estimator_grid(AgglomerativeClustering, n_clusters = 2:6, linkage = c("ward", "average")),
  estimator_grid(SpectralClustering, n_clusters = 2:6)
)
fit <- carve(X, estimator_param_grids = grids, n_resamples = 20, random_state = 0)
unique(estimator_results(fit)$method_label)
get_k(fit)
```

The grids set the numbers of clusters, so `n_clusters` is not needed.
Every argument with several values multiplies the configurations: this run
has `r nrow(estimator_results(fit))`. Without custom grids,
`estimator_param_grids = "light"` (the default) or `"full"` chooses a
preset; `carve()`'s help page lists what each holds.

## A clustering function of your own

Any function with the same shape works: the data matrix first, one row per
sample, then its parameters, returning one integer label per row and -1
for noise. If it has a `random_state` argument, CARVE passes it each
resample's seed. This one clusters on correlation distance:

```{r custom-estimator}
average_correlation <- function(X, n_clusters = 2L) {
  distances <- stats::as.dist(1 - stats::cor(t(X)))
  stats::cutree(stats::hclust(distances, method = "average"), k = n_clusters)
}
fit_custom <- carve(
  X, n_resamples = 20, random_state = 0,
  estimator_param_grids = list(estimator_grid(average_correlation, n_clusters = 2:6))
)
estimator_results(fit_custom)[, c("method_label", "n_clusters", "ari_stability")]
```

The method takes the name the function was passed under. An anonymous
function needs `estimator_grid(..., name = )`.

## Density-based clustering and noise

HDBSCAN takes a minimum cluster size instead of a number of clusters, and
labels the samples in sparse regions -1, for noise. It needs the dbscan
package, which CARVE suggests but does not install.

```{r hdbscan, eval = requireNamespace("dbscan", quietly = TRUE)}
fit_hdbscan <- carve(X, sweep = "min_cluster_size", sweep_values = c(5, 10),
                     n_resamples = 20, random_state = 0)
estimator_results(fit_hdbscan)[, c("min_cluster_size", "ari_stability", "ari_stability_se",
                                   "noise_fraction", "n_clusters_observed")]
get_sweep_value(fit_hdbscan)
```

A larger minimum size gives fewer clusters, so on this axis `"1se"` takes
the smallest size whose score is within one standard error of the best.
Size 5 scores more than one standard error below size 10, so the
selection is 10. `HDBSCAN()` can also be used in a grid, with
`estimator_grid(HDBSCAN, min_cluster_size = ...)`.

`noise_policy` sets what happens to the -1 labels in each resample.
`"drop"`, the default, leaves those samples out, so the consensus matrix,
the ARIs and the classifier treat them as not drawn. `"as_cluster"` keeps
-1 as an ordinary label, and `"singleton"` puts each noise sample in a
cluster of its own. `noise_fraction` is the share of the samples labeled
-1, averaged over the resamples. The same run with `"as_cluster"` draws
the same subsamples, so only the scores change:

```{r hdbscan-policy, eval = requireNamespace("dbscan", quietly = TRUE)}
fit_kept <- carve(X, sweep = "min_cluster_size", sweep_values = c(5, 10),
                  noise_policy = "as_cluster", n_resamples = 20, random_state = 0)
data.frame(min_cluster_size = c(5, 10),
           drop = estimator_results(fit_hdbscan)$ari_stability,
           as_cluster = estimator_results(fit_kept)$ari_stability)
get_sweep_value(fit_kept)
```

Scoring the noise samples as one more cluster lowers the stability at both
sizes here, and the selection stays at 10.

## The classifier

Generalizability uses a random forest of `n_trees` trees by default.
`classifier` takes another one: a function of `x_train`, `y_train` and
`x_test` that returns one predicted label per row of `x_test`. If it has
`n_threads` or `random_state` arguments, CARVE fills them in. A
five-nearest-neighbor classifier, with the FNN package that CARVE already
uses:

```{r classifier}
nearest_neighbors <- function(x_train, y_train, x_test) {
  FNN::knn(x_train, x_test, y_train, k = 5)
}
fit_knn <- carve(
  X, classifier = nearest_neighbors, n_resamples = 20, random_state = 0,
  estimator_param_grids = list(estimator_grid(KMeans, n_clusters = 2:6))
)
get_k(fit_knn, measure = "generalizability")
```

## Randomized preprocessing

Clustering often runs on transformed data: scaled, reduced with PCA, or
embedded with t-SNE or UMAP. With `randomize_preprocessing = TRUE`, each
resample draws a pipeline of one normalization option and one
dimensionality reduction option, fits it on its subsamples and clusters
the result, so a configuration that only works after one transform scores
lower. The classifier trains on the raw features. The transforms are
plain functions too:

```{r transforms}
dim(PCA(X, n_components = 2, random_state = 0))
round(colMeans(StandardScaler(X))[1:3], 10)
counts <- matrix(rpois(20, lambda = 5), nrow = 4)
Log1p(counts)
dim(TSNE(X, perplexity = 10, random_state = 0))
```

```{r umap, eval = requireNamespace("uwot", quietly = TRUE)}
dim(UMAP(X, n_neighbors = 10, random_state = 0))
```

`preprocessing_option()` pairs a transform with the values to draw from.
Without options, CARVE uses identity, standardization and, for data
without negative values, log1p, then identity, PCA, t-SNE and, with uwot
installed, UMAP. Here the options are smaller, to keep the run short:

```{r randomized}
fit_random <- carve(
  X, randomize_preprocessing = TRUE, n_resamples = 24, random_state = 0,
  estimator_param_grids = list(estimator_grid(KMeans, n_clusters = 2:6)),
  normalization_options = list(preprocessing_option(Identity), preprocessing_option(StandardScaler)),
  dim_reduction_options = list(preprocessing_option(Identity), preprocessing_option(PCA, n_components = c(2, 5)))
)
head(preprocessing_results(fit_random)[, c("pipeline", "n_clusters", "n_resamples", "ari_stability")])
```

```{r pipeline-plot, fig.width = 7}
plot_metric_by_pipeline(fit_random)
```

`preprocessing_results()` splits each configuration's scores by the
pipeline its resamples used, and `plot_metric_by_pipeline()` draws one
line per pipeline for the selected method. `preprocessing_pipelines()`
returns the pipelines by label, and `pipeline_from_spec()` turns one into
a function that applies it to data, for example to embed all the samples:

```{r pipelines}
pipelines <- preprocessing_pipelines(fit_random)
names(pipelines)
embed <- pipeline_from_spec(pipelines[[1]], random_state = 0)
dim(embed(X))
```

## Reference labels

The cluster numbers `get_labels()` returns are arbitrary. `reference_labels`
renames the clusters to match another labeling with the same number of
clusters, by maximum overlap. Given to `carve()`, it applies to every
`get_labels()` call on the fit:

```{r reference}
fit_reference <- carve(
  X, reference_labels = truth, n_resamples = 20, random_state = 0,
  estimator_param_grids = list(estimator_grid(KMeans, n_clusters = 2:6))
)
table(get_labels(fit_reference), truth)
```

`get_labels()` keeps nothing between calls. To keep the names of one
call's clusters in a later call, pass its labels. The reference needs as many
clusters as the new cut, so this call fixes `k`:

```{r reference-call}
first <- get_labels(fit)
table(get_labels(fit, measure = "generalizability", k = 4, reference_labels = first), first)
```

## Parallel runs and memory

`n_jobs` is a core budget. The resamples of each configuration are spread
over up to `n_jobs` worker processes, and each worker's random forest gets
the cores divided by the number of workers. `n_jobs = -1` uses every core.
The results do not depend on it:

```{r parallel}
grid <- list(estimator_grid(KMeans, n_clusters = 2:6))
fit_two <- carve(X, estimator_param_grids = grid, n_jobs = 2, n_resamples = 20, random_state = 0)
fit_one <- carve(X, estimator_param_grids = grid, n_resamples = 20, random_state = 0,
                 BPPARAM = BiocParallel::SerialParam())
all.equal(estimator_results(fit_two), estimator_results(fit_one))
```

`BPPARAM` chooses the BiocParallel backend; its worker count then replaces
the one `n_jobs` would set. `show_progress = TRUE` shows a progress bar
over the configurations.

Each worker holds its own copy of the data. The fit keeps two consensus
matrices of `n` by `n` doubles per configuration, `8 * n^2` bytes each, so
memory grows with the square of the number of samples. `verbose = 2`
prints the projected total in its header, as the `consensus memory` line.
It reads 0.00 GB here because the data have only 240 samples:

```{r verbose}
fit_verbose <- carve(X, estimator_param_grids = list(estimator_grid(KMeans, n_clusters = 2:3)),
                     n_resamples = 5, random_state = 0, verbose = 2)
```

`vignette("single-cell", package = "CARVE")` shows how anchors keep that
memory bounded on large data.

## Session information

```{r session}
sessionInfo()
```
