---
title: "Getting started with CARVE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CARVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 90
)
set.seed(1)
```

CARVE validates clustering by repeatedly subsampling the data, running
candidate `(estimator, k)` configurations on each subsample, and
summarising the per-resample agreement with two complementary metric
families:

* **Stability** — how reproducible are the cluster assignments
  across resamples? (`ari_stability`, `consensus_pac_stability`,
  `consensus_gini_stability`, `consensus_ce_stability`)
* **Generalizability** — how well do held-out points get classified
  back into the in-sample clusters by a `ranger` random forest?
  (`accuracy_generalizability`, `ari_generalizability`)

A selection rule (`max`, `1se`, or `quantile`) then turns the
metric-vs-`k` curve into a single `(estimator, k)` recommendation.

## A toy dataset

Three well-separated 2-D Gaussian blobs, 50 points each:

```{r toy-data}
library(CARVE)

centers <- rbind(c(0, 0), c(6, 0), c(3, 5))
X <- do.call(rbind, lapply(seq_len(nrow(centers)), function(i) {
  matrix(stats::rnorm(50 * 2, 0, 0.4), ncol = 2) +
    matrix(centers[i, ], 50, 2, byrow = TRUE)
}))
y_true <- rep(seq_len(nrow(centers)), each = 50)

dim(X)
```

## Fitting CARVE

The R6 form gives reference-semantics mutation that mirrors the
Python class:

```{r fit}
fit <- CARVE$new(
  n_clusters = 2:5,
  n_resamples = 30L,
  subsample_ratio = 0.8,
  estimator_param_grids = list(
    list(type = "kmeans", grid = list(n_clusters = 2:5))
  ),
  random_state = 1L,
  verbose = 0L
)
fit$fit(X)
```

The S3 form `carve(X, ...)` is equivalent and friendlier inside
pipelines:

```{r s3-equivalent, eval = FALSE}
fit <- carve(X, n_clusters = 2:5, n_resamples = 30L,
             subsample_ratio = 0.8, random_state = 1L)
```

## Inspecting the metrics

After `$fit()` the per-configuration metrics live in
`fit$estimator_results_`:

```{r results-table}
res <- fit$estimator_results_
head(res[, c("estimator", "n_clusters",
             "ari_stability", "consensus_pac_stability",
             "ari_generalizability", "accuracy_generalizability")])
```

## Picking a `k`

```{r get-k}
fit$get_k(measure = "stability", rule = "max")
fit$get_k(measure = "generalizability", rule = "1se")
```

The labels for the chosen solution are produced by re-clustering the
selected stability consensus matrix with average-linkage hierarchical
clustering, mirroring the Python implementation:

```{r get-labels}
labels <- fit$get_labels(measure = "stability", rule = "max")
table(true = y_true, carve = labels)
```

## Plotting

Every plot method returns a `ggplot` (or, for the consensus heatmap,
a `patchwork`):

```{r plot-metric}
fit$plot_metric_over_n_clusters(measure = "stability", rule = "max")
```

```{r plot-consensus}
fit$plot_consensus_matrix(measure = "stability", rule = "max")
```

```{r plot-cluster-summary}
fit$plot_cluster_violin(source = "gini", measure = "stability",
                        rule = "max", stripplot = TRUE)
fit$plot_cluster_scatter(source = "gini", measure = "stability",
                         rule = "max")
```

Combine plots with `patchwork`:

```{r patchwork-combo, fig.width = 9, fig.height = 4}
library(patchwork)
fit$plot_metric_over_n_clusters(measure = "stability", rule = "max") +
  fit$plot_cluster_scatter(source = "gini", measure = "stability",
                           rule = "max")
```

## Saving and reloading

```{r save-load, eval = FALSE}
path <- tempfile(fileext = ".carve")
fit$save(path)
again <- CARVE$load(path)
identical(fit$get_labels(), again$get_labels())
```

By default `save()` strips `X_` so the on-disk file stays small;
pass `include_data = TRUE` to keep it.

## Going further

* `vignette("seurat-workflow", package = "CARVE")` — end-to-end
  Seurat pipeline.
* `vignette("cross-validation", package = "CARVE")` — Python ↔ R
  numerical equivalence on fixed seeds.
