---
title: "CARVE in a Seurat workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CARVE in a Seurat workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_seurat <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("SeuratObject", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 90,
  eval = has_seurat
)
```

```{r seurat-missing, eval = !has_seurat, echo = FALSE, results = "asis"}
cat("> **Note.** This vignette is rendered without `Seurat` installed,",
    "so chunks below are not evaluated. Install `Seurat` from CRAN to",
    "run the workflow end-to-end:\n>\n> ",
    "```r\n> install.packages(\"Seurat\")\n> ```\n")
```

`Seurat` is in `Suggests`, so CARVE works with Seurat objects but
never imports the package at load time. The integration has three
entry points:

1. `carve(seurat_obj, ...)` — S3 dispatch returning a fitted `CARVE`.
2. `RunCARVE(seurat_obj, ...)` — verb-style alias that stores the
   fit on `obj@misc$CARVE`.
3. `AddCarveLabels(seurat_obj, measure, rule)` — writes the selected
   labels into `obj@meta.data` and (optionally) sets the active
   ident.

Throughout we use the bundled `pbmc_small` dataset (80 cells × 230
features), which is small enough to run in a vignette in seconds.

## Fitting on a PCA reduction

```{r seurat-fit}
suppressPackageStartupMessages(library(Seurat))
library(CARVE)

data("pbmc_small")
pbmc_small

pbmc_small <- RunCARVE(
  pbmc_small,
  reduction = "pca",
  n_dims = 10,
  n_clusters = 2:6,
  n_resamples = 20L,
  subsample_ratio = 0.8,
  estimator_param_grids = list(
    list(type = "kmeans", grid = list(n_clusters = 2:6))
  ),
  random_state = 1L
)

fit <- pbmc_small@misc$CARVE
fit$get_k(measure = "stability", rule = "max")
```

`RunCARVE()` forwards `reduction = "pca"` through the
`carve.Seurat` extractor, which in turn pulls the embedding via
`Seurat::Embeddings()` and trims to the first `n_dims` columns.

## Adding labels back to the object

```{r seurat-labels}
pbmc_small <- AddCarveLabels(
  pbmc_small,
  measure = "stability",
  rule = "max",
  label_col = "carve_labels"
)

table(pbmc_small$carve_labels)
```

`AddCarveLabels()` writes a factor column to `meta.data` and (by
default) sets it as the active ident. Pass `set_ident = FALSE` to
keep the existing ident.

## Visualising on UMAP / PCA

```{r seurat-dimplot}
DimPlot(pbmc_small, reduction = "pca", group.by = "carve_labels")
```

CARVE's own plot methods still work — they show resample-level
diagnostics, not per-cell embeddings:

```{r seurat-carve-plots}
fit$plot_metric_over_n_clusters(measure = "stability", rule = "max")
fit$plot_consensus_matrix(measure = "stability", rule = "max")
```

## Working with assay data directly

`reduction` is optional. To run CARVE on a normalised expression
matrix instead, leave `reduction = NULL` and pick an `assay` /
`slot`:

```{r seurat-assay, eval = FALSE}
pbmc_small <- RunCARVE(
  pbmc_small,
  assay = "RNA", slot = "data",
  features = VariableFeatures(pbmc_small),
  n_clusters = 2:6, n_resamples = 20L, random_state = 1L
)
```

Seurat v5 renamed `slot` to `layer`; `RunCARVE()` dispatches on
`packageVersion("Seurat")` so the same `slot = ...` argument works
on both v4 and v5.

## SingleCellExperiment

A symmetric S3 method exists for `SingleCellExperiment`:

```{r sce, eval = FALSE}
library(SingleCellExperiment)
sce <- SingleCellExperiment(
  assays = list(logcounts = t(Embeddings(pbmc_small, "pca")))
)
fit_sce <- carve(sce, assay = "logcounts",
                 n_clusters = 2:6, n_resamples = 20L,
                 random_state = 1L)
```

## Saving the fit

CARVE saves cleanly with `saveRDS`-style semantics; the embedded
matrix `X_` is stripped by default to keep the file small.

```{r seurat-save, eval = FALSE}
fit$save("pbmc_small_carve.rds")
fit2 <- CARVE$load("pbmc_small_carve.rds")
```
