---
title: "CARVE for single-cell data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CARVE for single-cell data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

Single-cell data are usually clustered with a community detection method,
Leiden or Louvain, on a nearest-neighbor graph of the cells' principal
components. These methods take a resolution instead of a number of
clusters. CARVE compares resolutions the way it compares numbers of
clusters, and records, for each resolution, the mean number of clusters it gives over
the resamples. This vignette fits CARVE to a
SingleCellExperiment and a Seurat object.
`vignette("CARVE", package = "CARVE")` explains the scores and the
selection rules.

The data are `pbmc3k_subset`: 1,000 cells of 10x Genomics' PBMC 3k data
set, with 30 principal components, a UMAP and the clusters of Seurat's
tutorial, all computed on the full data.

```{r sce}
library(CARVE)

cells <- rownames(pbmc3k_subset$pca)
sce <- SingleCellExperiment::SingleCellExperiment(
  reducedDims = list(PCA = pbmc3k_subset$pca, UMAP = pbmc3k_subset$umap),
  colData = S4Vectors::DataFrame(seurat_clusters = pbmc3k_subset$seurat_clusters, row.names = cells)
)
sce
```

## Comparing resolutions

```{r run}
sce <- run_carve(sce, resolution = c(0.25, 0.5, 1, 1.5), n_resamples = 20)
```

`run_carve()` reads the cells from the `"PCA"` reduced dimensions, fits
CARVE and writes the result into the object. `reduction`, `assay` and
`n_dims` choose other input; without a `"PCA"` entry it reads the
`"logcounts"` assay. With `resolution` given, CARVE compares Leiden and
Louvain clustering, each on a graph of the 15 nearest neighbors, at every
resolution. Each subsample is clustered on a graph with the neighbor count
scaled to its size, so that the graph covers the same neighborhood. This
run uses 20 resamples to keep the vignette fast; use the default, 100, for
an analysis.

The selected configuration's results go into the object:

```{r stored}
head(SummarizedExperiment::colData(sce))
names(S4Vectors::metadata(sce)$carve)
S4Vectors::metadata(sce)$carve$params$selected_resolution
table(CARVE = sce$carve, Seurat = sce$seurat_clusters)
```

The column `carve` holds the labels, as a factor. `carve_stability` and
`carve_stability_ce` hold each cell's stability in its Gini and
cross-entropy forms, and `carve_generalizability` the share of the draws
holding the cell out in which the classifier predicted its cluster.
`metadata(sce)$carve` holds the run's parameters, the results table and
the selected configuration's consensus matrix. The `key` argument changes
the name `carve` in all of them, so the results of several runs can sit in
one object.

CARVE's `r nlevels(sce$carve)` clusters are coarser than the
`r nlevels(sce$seurat_clusters)` of Seurat's tutorial. In the table, its
cluster 1 holds most cells of Seurat's clusters 0 and 2, and its cluster 2
most cells of clusters 4 and 6.

## Plots of the object

Every plot function also takes the object and reads what `run_carve()`
stored:

```{r sce-metric, fig.width = 7}
plot_metric_over_n_clusters(sce)
plot_n_clusters_over_sweep(sce)
```

The first plot has the resolution on the x axis. The second shows how
many clusters each method found at each resolution, averaged over the
resamples, with standard error bars.

```{r sce-cells, fig.width = 6, fig.height = 4.5}
plot_cluster_scatter(sce, basis = "UMAP")
plot_cluster_violin(sce)
```

`basis` names the reduced dimensions to draw. Cells with low stability are
drawn larger and more opaque.

## Keeping the fit

`run_carve()` stores only the selected configuration, because a fit holds
the consensus matrices of all of them. To look at the others, fit with
`carve()` and keep the fit. With the same arguments and seed it gives the
same result, and `attach_results()` writes it into the object as
`run_carve()` does:

```{r fit}
fit <- carve(sce, resolution = c(0.25, 0.5, 1, 1.5), n_resamples = 20, random_state = 0)
get_sweep_value(fit)
get_k(fit)
identical(attach_results(sce, fit)$carve, sce$carve)
```

On a resolution sweep, `get_sweep_value()` returns the selected
resolution, and `get_k()` the number of clusters at it, rounded from the
mean over the resamples. `sweep_value` asks for another resolution, and
`consensus_k` cuts the selected consensus matrix into another number of
clusters:

```{r other-resolution}
table(get_labels(fit, sweep_value = 1))
table(get_labels(fit, consensus_k = 4))
sce <- attach_results(sce, fit, sweep_value = 1, key = "carve_res1")
table(sce$carve_res1)
```

## Ambiguous cells

`noise_labels = TRUE` labels -1 the cells whose stability is low compared
with the rest: those strictly below the `noise_quantile` of the selected
configuration's per-cell scores and more than 0.05 below their median.
`noise_score` picks the score.

```{r noise}
table(get_labels(fit, noise_labels = TRUE, noise_quantile = 0.05))
```

If fewer cells pass both tests than the quantile asks for, a warning gives
the counts.

## Larger data sets

A consensus matrix has one entry per pair of cells, and the fit keeps two
per configuration, `8 * n^2` bytes each. Above `anchor_threshold` cells,
5,000 by default, CARVE draws `consensus_anchors` anchor cells once and
keeps the consensus over them only. Every cell still gets its scores, and
`get_labels()` labels the cells that are not anchors with the classifier.
Setting `consensus_anchors` anchors a smaller run too, with a warning,
which is how it is shown here:

```{r anchored}
anchored <- carve(sce, resolution = c(0.5, 1), n_resamples = 20,
                  consensus_anchors = 300, random_state = 0)
length(consensus_anchors(anchored))
dim(consensus_matrix(anchored, estimator_results(anchored)$config_id[1]))
table(get_labels(anchored))
```

## Seurat objects

The same functions take a Seurat object. Its default input is the `"pca"`
reduction, then the `"data"` layer of the default assay; `assay` names a
Seurat assay, whose `"data"` layer is clustered. The labels and scores go
into the cell metadata, and the record into the `misc` slot.
`SeuratObject::pbmc_small` has 80 cells:

```{r seurat, eval = requireNamespace("SeuratObject", quietly = TRUE)}
pbmc <- SeuratObject::pbmc_small
pbmc <- run_carve(pbmc, resolution = c(0.5, 1, 2), n_resamples = 20)
head(pbmc@meta.data[, c("carve", "carve_stability", "carve_generalizability")])
names(pbmc@misc$carve)
```

```{r seurat-plot, eval = requireNamespace("SeuratObject", quietly = TRUE), fig.width = 6, fig.height = 4.5}
plot_cluster_scatter(pbmc, basis = "tsne")
```

## Session information

```{r session}
sessionInfo()
```
