---
output:
  github_document:
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Edit the .Rmd and run -->
<!-- devtools::build_readme() to regenerate. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# CARVE

CARVE (cluster analysis with resampling for validation and exploration)
chooses the number of clusters, and the clustering method, by resampling.
Each candidate configuration clusters many subsamples of the data and is
scored for stability, the agreement between the clusterings of two
overlapping subsamples, and for generalizability, how well a classifier
trained on one subsample predicts the clusters of the held-out samples.
CARVE compares numbers of clusters, Leiden and Louvain resolutions, or
HDBSCAN's minimum cluster size, on a matrix or on a SingleCellExperiment
or Seurat object.

This is the R package. The Python package, carve-validate, is in the same
repository; both compute the same scores with the same defaults.

## Installation

```r
# install.packages("remotes")
remotes::install_github("DataSlingers/CARVE", subdir = "carve-r")
```

Add `build_vignettes = TRUE` to build the vignettes. HDBSCAN needs the
dbscan package, UMAP the uwot package and Seurat objects the SeuratObject
package; CARVE suggests them but does not install them.

## Example

```{r example}
library(CARVE)

set.seed(1)
centers <- rbind(c(0, 0), c(4, 0), c(2, 3.5), c(6, 3.5))
X <- centers[rep(1:4, each = 50), ] + matrix(rnorm(400, sd = 0.8), ncol = 2)

fit <- carve(X, n_clusters = 2:8, n_resamples = 30, random_state = 1)
get_k(fit)
table(get_labels(fit))
```

`plot_metric_over_n_clusters(fit)` draws the scores against the number of
clusters, and `estimator_results(fit)` holds them as a table.

## Documentation

- `vignette("CARVE", package = "CARVE")`: the scores, the selection rules,
  the results and the plots, on simulated data.
- `vignette("single-cell", package = "CARVE")`: SingleCellExperiment and
  Seurat objects, Leiden and Louvain resolutions, and large data sets.
- `vignette("customizing", package = "CARVE")`: methods and parameters,
  HDBSCAN, the classifier, randomized preprocessing, reference labels and
  parallel runs.
- `vignette("python-users", package = "CARVE")`: the Python names and
  their R equivalents, and where the results differ.
- `notebooks/R_Tutorial.Rmd` in the repository goes through every function
  on full-size data, and `notebooks/R_Tutorial.html` is its rendered
  output.

## Citation

If you use CARVE, please cite:

> Wycik, K. R., Tang, T. M., Zikry, T. M., & Allen, G. I. (2026). CARVE:
> Cluster Analysis with Resampling for Validation and Exploration. Zenodo.
> https://doi.org/10.5281/zenodo.20448965

```bibtex
@software{wycik2026carve,
  author    = {Wycik, Kai R. and Tang, Tiffany M. and Zikry, Tarek M. and Allen, Genevera I.},
  title     = {{CARVE}: Cluster Analysis with Resampling for Validation and Exploration},
  year      = {2026},
  publisher = {Zenodo},
  doi       = {10.5281/zenodo.20448965},
  url       = {https://doi.org/10.5281/zenodo.20448965}
}
```

## License

MIT. The HDBSCAN cluster selection is ported from scikit-learn and keeps
its BSD 3-Clause notice; see `inst/COPYRIGHTS`.
