---
title: "CARVE tutorial for R"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
tutorial_start <- Sys.time()
```

This tutorial goes through the CARVE R package, on simulated data in
part 1 and on the PBMC 3k single-cell data set in part 2. It follows the
Python notebooks `Tutorial.ipynb` and `Resolution_Tutorial.ipynb` in this
folder and fits at full size, with the default 100 resamples except in the
randomized preprocessing fits, which use 60. The
package's vignettes explain the same steps on smaller data:
`vignette("CARVE", package = "CARVE")` for the scores and the selection
rules, `vignette("single-cell", package = "CARVE")`,
`vignette("customizing", package = "CARVE")` and
`vignette("python-users", package = "CARVE")`.

Part 2 needs the Seurat, SeuratObject, dbscan and mclust packages, and
downloads 7 MB from 10x Genomics into `data/` on its first run.

```{r packages}
library(CARVE)

n_jobs <- max(1L, parallel::detectCores() - 1L)
n_jobs
```

`n_jobs` is the number of cores each fit may use.

# Part 1: simulated data

## Quick start

Two pairs of interleaved half-moons, 1,000 points in four groups. The
groups are curved, so they are not convex clusters.

```{r moons, fig.width = 6, fig.height = 4}
make_moons <- function(n, noise) {
  n_outer <- n %/% 2
  n_inner <- n - n_outer
  t_outer <- seq(0, pi, length.out = n_outer)
  t_inner <- seq(0, pi, length.out = n_inner)
  X <- rbind(cbind(cos(t_outer), sin(t_outer)),
             cbind(1 - cos(t_inner), 1 - sin(t_inner) - 0.5))
  X + matrix(rnorm(2 * n, sd = noise), ncol = 2)
}

set.seed(42)
moons <- rbind(make_moons(500, noise = 0.12),
               make_moons(500, noise = 0.10) + matrix(c(2, 1.5), 500, 2, byrow = TRUE))
moons_truth <- rep(1:4, each = 250)
plot(moons, col = moons_truth, pch = 19, cex = 0.5, asp = 1, xlab = "", ylab = "")
```

```{r moons-fit}
moons_grids <- list(
  estimator_grid(AgglomerativeClustering, n_clusters = 2:6, linkage = "ward"),
  estimator_grid(SpectralClustering, n_clusters = 2:6, affinity = "self_tuning")
)
fit_moons <- carve(moons, estimator_param_grids = moons_grids,
                   n_jobs = n_jobs, random_state = 42)
fit_moons
get_k(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
moons_labels <- get_labels(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
table(moons_labels, moons_truth)
moons_ari <- mclust::adjustedRandIndex(moons_labels, moons_truth)
moons_ari
```

This fit uses two methods, ward-linkage agglomerative clustering and
self-tuning spectral clustering, each with 2 to 6 clusters. The selection
scores each configuration by generalizability, takes the largest number of
clusters within one standard error of the best (`rule = "1se"`), and
leaves two clusters out (`not_two = TRUE`). The two pairs of moons are far
apart, so a split into two is easy to reproduce, and the default selection
returns `r get_k(fit_moons)`; the section on two clusters below shows the
same effect on other data. With the selection above, CARVE selects
`r get_k(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)`
clusters, and the labels have an ARI of `r round(moons_ari, 2)` with the
four true groups. The table shows that the first pair of moons is one
cluster, while the second pair is split into two clusters that follow its
moons except for `r sum(moons_labels == 2 & moons_truth == 3)` points. The two moons of the first pair are
not separated.

```{r moons-plots}
plot_metric_over_n_clusters(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
plot_cluster_scatter(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
```

## How CARVE scores a configuration

A configuration is a clustering method, its parameters and a value of the
swept parameter, here the number of clusters. For each of `n_resamples`
resamples, CARVE draws two overlapping subsamples of `subsample_ratio` of
the samples (0.618 by default) and clusters both. Stability is the
adjusted Rand index (ARI) between the two clusterings on the samples they
share. For generalizability, a random forest trained on one subsample's
clusters predicts the clusters of the samples held out of it, and the ARI
compares the prediction with the clusters those samples were given. After
the resamples, CARVE builds each configuration's consensus matrix, the
share of the subsamples drawing two samples in which they fell in the
same cluster, and computes per-sample scores from it.

```{r results}
results <- estimator_results(fit_moons)
names(results)
results[results$n_clusters == 4,
        c("method_label", "ari_stability", "ari_generalizability",
          "consensus_pac_stability", "consensus_gini_stability", "accuracy_generalizability")]
```

`ari_stability` and `ari_generalizability` are means over the resamples,
with standard errors (`_se`) and 5th and 95th percentiles (`_lower`,
`_upper`). `consensus_pac_stability` is one minus the proportion of
ambiguous clustering, the share of consensus entries between 0.05 and 0.95,
so higher is better, as for every score in the table.
`consensus_gini_stability` and `consensus_ce_stability` average the
per-sample stability scores, and `accuracy_generalizability` the
per-sample accuracy of the classifier on held-out samples.

## Plots

The eight plots return ggplot objects. The metric plot draws one line per
method over the swept values, and the dashed line marks the selection. The
quick start passed the selection arguments to it. With the defaults
(stability, the 1se rule, two clusters allowed) the dashed line is at
`r get_k(fit_moons)` clusters:

```{r plot-metric}
plot_metric_over_n_clusters(fit_moons)
```

`plot_n_clusters_over_sweep()` draws the number of clusters each method
returned, averaged over the resamples. It needs a sweep over a parameter
other than `n_clusters`, such as a resolution, where the count is an
outcome; on an `n_clusters` sweep the count is the swept value itself and
the function stops with an error saying so. Part 2 uses it.

The consensus heatmap orders the samples by cluster. `k` shows another
number of clusters:

```{r plot-consensus, fig.width = 5.5, fig.height = 5}
plot_consensus_matrix(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
plot_consensus_matrix(fit_moons, k = 6)
```

The per-sample plots show the stability of each sample (`source = "gini"`
or `"ce"`) or the share of the draws holding it out in which the
classifier predicted its cluster (`source = "accuracy"`):

```{r plot-per-sample}
plot_cluster_boxplot(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
plot_cluster_violin(fit_moons, source = "accuracy", measure = "generalizability", rule = "1se", not_two = TRUE)
plot_diagnostic_scatter(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
```

`plot_metric_by_pipeline()`, the eighth, is in the section on randomized
preprocessing below.

## Selecting a configuration and labeling the samples

```{r selecting}
c(default = get_k(fit_moons),
  generalizability = get_k(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE))
get_sweep_value(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
moons_estimator <- get_estimator(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)
attributes(moons_estimator)[c("estimator", "params")]
table(get_labels(fit_moons, k = 2), moons_truth)
```

`get_k(fit_moons)` with the defaults returns `r get_k(fit_moons)`, and
the selection of the quick start returns
`r get_k(fit_moons, measure = "generalizability", rule = "1se", not_two = TRUE)`.
`get_estimator()` returns the selected method as a function of
the data and a seed, and `get_labels(k = 2)` cuts the best configuration
at two clusters instead of the selected one. Labels run from 1 to the number of clusters.

## Easy and hard data

Three well-separated groups, and five that overlap:

```{r easy-hard}
make_blobs <- function(n_per, centers, sd) {
  truth <- rep(seq_len(nrow(centers)), each = n_per)
  noise <- matrix(rnorm(length(truth) * ncol(centers), sd = sd), ncol = ncol(centers))
  list(X = centers[truth, ] + noise, truth = truth)
}

set.seed(0)
easy <- make_blobs(100, rbind(c(0, 0), c(8, 0), c(4, 7)), sd = 1)
hard <- make_blobs(100, rbind(c(0, 0), c(3, 0), c(1.5, 2.6), c(4.5, 2.6), c(3, 5.2)), sd = 1.2)
fit_easy <- carve(easy$X, n_clusters = 2:8, n_jobs = n_jobs, random_state = 0)
fit_hard <- carve(hard$X, n_clusters = 2:8, n_jobs = n_jobs, random_state = 0)
c(easy = get_k(fit_easy), hard = get_k(fit_hard))
```

```{r easy-hard-plots}
plot_metric_over_n_clusters(fit_easy) + ggplot2::labs(title = "Three separated groups")
plot_metric_over_n_clusters(fit_hard) + ggplot2::labs(title = "Five overlapping groups")
```

On the easy data the scores of all three methods peak at the true three
groups, at an ARI near 1. On the hard data the scores are lower and the
methods disagree, and the best score, from k-means, is at two clusters,
not at the five true groups.

## The selection rules

The same hard fit under each measure and rule:

```{r rules}
measures <- c("stability", "generalizability", "average")
rules <- c("max", "1se", "quantile")
sapply(rules, function(rule) {
  sapply(measures, function(measure) get_k(fit_hard, measure = measure, rule = rule))
})
```

`"max"` takes the highest mean score. `"1se"` takes the largest number of
clusters whose score is within one standard error of the highest, and
`"quantile"` the largest within the 5th to 95th percentiles of the best
configuration's resample scores, so both lean toward finer clusterings
when scores are close. On this fit `"max"` and `"1se"` give
`r paste(unique(c(sapply(measures, function(m) c(get_k(fit_hard, measure = m, rule = "max"), get_k(fit_hard, measure = m, rule = "1se"))))), collapse = ", ")`
clusters, whichever measure is used. `"quantile"` gives
`r get_k(fit_hard, measure = "stability", rule = "quantile")` for stability
and `r get_k(fit_hard, measure = "generalizability", rule = "quantile")`
for generalizability, so the two scores can point to different numbers of
clusters when the rule is looser.

## Two clusters

Four groups in a row can be split into two halves the same way in every
resample, so two clusters can score as well as four:

```{r not-two}
set.seed(1)
line <- make_blobs(75, cbind(c(0, 3, 6, 9), 0), sd = 0.7)
fit_line <- carve(line$X, n_clusters = 2:8, n_jobs = n_jobs, random_state = 1)
c(max = get_k(fit_line, rule = "max"),
  max_not_two = get_k(fit_line, rule = "max", not_two = TRUE),
  one_se = get_k(fit_line))
```

```{r not-two-plots, fig.width = 5.5, fig.height = 5}
plot_consensus_matrix(fit_line, k = 2)
plot_consensus_matrix(fit_line, rule = "max", not_two = TRUE)
```

The two-cluster consensus matrix looks clean, but it says little about
the four groups. `not_two = TRUE` leaves two clusters out of the
selection.

## Custom grids

`estimator_grid()` sets the methods and the values of their parameters:

```{r grids}
grids <- list(
  estimator_grid(KMeans, n_clusters = 2:8, n_init = 10),
  estimator_grid(AgglomerativeClustering, n_clusters = 2:8, linkage = c("ward", "single")),
  estimator_grid(SpectralClustering, n_clusters = 2:8, n_neighbors = c(5, 15))
)
fit_grid <- carve(moons, estimator_param_grids = grids, n_jobs = n_jobs, random_state = 42)
unique(estimator_results(fit_grid)$method_label)
get_k(fit_grid)
plot_metric_over_n_clusters(fit_grid)
```

## Randomized preprocessing

k-means does not follow the curved shapes of the moons in their own
coordinates. This section shows how to set up a randomized run and how to
read its results. With `randomize_preprocessing = TRUE`, each resample draws a normalization and
a dimensionality reduction and clusters the transformed subsamples; the
classifier still trains on the raw coordinates.

```{r randomized}
fit_random <- carve(
  moons, randomize_preprocessing = TRUE, n_resamples = 60, n_jobs = n_jobs, random_state = 42,
  estimator_param_grids = list(estimator_grid(KMeans, n_clusters = 2:8)),
  normalization_options = list(preprocessing_option(Identity), preprocessing_option(StandardScaler)),
  dim_reduction_options = list(preprocessing_option(Identity), preprocessing_option(TSNE, perplexity = c(30, 50)))
)
plot_metric_by_pipeline(fit_random)
pipeline_results <- preprocessing_results(fit_random)
pipeline_results[pipeline_results$n_clusters == 4,
                 c("pipeline", "n_resamples", "ari_stability", "ari_generalizability")]
```

`preprocessing_results()` splits each configuration's scores by pipeline.
This run uses 60 resamples, because every resample runs t-SNE on its
subsamples, so each pipeline's scores rest on
`r min(pipeline_results$n_resamples)` to `r max(pipeline_results$n_resamples)`
resamples (`n_resamples` in the table) and are noisy. At four clusters the pipelines
with t-SNE do not score higher than the identity pipelines on these data.

## Saving a fit

```{r save}
path <- file.path(tempdir(), "fit_moons.rds")
saveRDS(fit_moons, path)
reloaded <- readRDS(path)
identical(get_labels(reloaded, measure = "generalizability", rule = "1se", not_two = TRUE), moons_labels)
round(file.size(path) / 2^20, 1)
```

The size is in MB. A fit keeps the data, so the scatter plots work on a
fit read back from disk; Python's `save()` leaves the data out unless
`include_data=True`.

## Reference labels

Cluster numbers are arbitrary. `reference_labels` renames the clusters to
match another labeling with the same number of clusters, here the truth:

```{r reference}
table(get_labels(fit_moons, k = 4, reference_labels = moons_truth), moons_truth)
```

`carve(reference_labels = )` sets the labeling for every `get_labels()`
call on the fit.

## Parallel runs

A run clusters every configuration on every resample, so its cost grows
with the number of configurations times `n_resamples`. `n_jobs` spreads
each configuration's resamples over worker processes; the results do not
depend on it.

```{r timing}
blob_grid <- list(estimator_grid(KMeans, n_clusters = 2:8))
time_serial <- system.time(fit_serial <- carve(hard$X, estimator_param_grids = blob_grid, random_state = 0))
time_workers <- system.time(fit_workers <- carve(hard$X, estimator_param_grids = blob_grid, n_jobs = n_jobs, random_state = 0))
c(serial = time_serial[["elapsed"]], workers = time_workers[["elapsed"]])
all.equal(estimator_results(fit_serial), estimator_results(fit_workers))
```

Starting the workers has a cost of its own, so the gain grows with the
work in each resample.

# Part 2: PBMC 3k

Single-cell data are usually clustered with Leiden or Louvain community
detection on a nearest-neighbor graph of principal components. These
methods take a resolution, and CARVE compares resolutions the way it
compares numbers of clusters; the number of clusters becomes an outcome.

## Data and the Seurat workflow

The 10x Genomics PBMC 3k data, processed with the steps of Seurat's
tutorial:

```{r pbmc-download}
url <- "https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz"
dir.create("data", showWarnings = FALSE)
archive <- file.path("data", basename(url))
if (!file.exists(archive)) {
  utils::download.file(url, archive, mode = "wb")
}
matrix_dir <- file.path("data", "filtered_gene_bc_matrices", "hg19")
if (!dir.exists(matrix_dir)) {
  utils::untar(archive, exdir = "data")
}
```

```{r pbmc-seurat, message = FALSE, warning = FALSE}
pbmc <- SeuratObject::CreateSeuratObject(
  counts = Seurat::Read10X(matrix_dir), project = "pbmc3k", min.cells = 3, min.features = 200
)
pbmc[["percent.mt"]] <- Seurat::PercentageFeatureSet(pbmc, pattern = "^MT-")
keep <- pbmc$nFeature_RNA > 200 & pbmc$nFeature_RNA < 2500 & pbmc$percent.mt < 5
pbmc <- subset(pbmc, cells = colnames(pbmc)[keep])
pbmc <- Seurat::NormalizeData(pbmc, verbose = FALSE)
pbmc <- Seurat::FindVariableFeatures(pbmc, nfeatures = 2000, verbose = FALSE)
pbmc <- Seurat::ScaleData(pbmc, verbose = FALSE)
pbmc <- Seurat::RunPCA(pbmc, npcs = 30, verbose = FALSE)
pbmc <- Seurat::FindNeighbors(pbmc, dims = 1:10, verbose = FALSE)
pbmc <- Seurat::FindClusters(pbmc, resolution = 0.5, verbose = FALSE)
pbmc <- Seurat::RunUMAP(pbmc, dims = 1:10, verbose = FALSE)
pbmc
```

The chunk hides Seurat's own messages and warnings, such as its notice
about the UMAP method; CARVE's output is shown everywhere else.

## Comparing resolutions

```{r pbmc-fit}
resolutions <- c(0.25, 0.5, 0.75, 1, 1.5)
fit_pbmc <- carve(pbmc, resolution = resolutions, n_dims = 10, n_jobs = n_jobs, random_state = 0)
fit_pbmc
```

`carve()` reads the first 10 principal components of the `"pca"`
reduction, the ones Seurat's graph used. With `resolution` given, it
compares Leiden and Louvain at each resolution, each on a graph of the 15
nearest neighbors scaled to the subsample's size.

```{r pbmc-sweep}
sweep_spec(fit_pbmc)
get_sweep_value(fit_pbmc)
get_k(fit_pbmc)
```

`sweep_spec()` describes the sweep: the parameter, its values, and that
larger values give more clusters. `get_sweep_value()` returns the selected
resolution, and `get_k()` the number of clusters at it, rounded from the
mean over the resamples.

```{r pbmc-curves}
plot_metric_over_n_clusters(fit_pbmc)
plot_n_clusters_over_sweep(fit_pbmc)
```

The metric plot has the resolution on its x axis, one line per method.
The second plot shows how many clusters each method found at each
resolution.

```{r pbmc-consensus, fig.width = 5.5, fig.height = 5}
plot_consensus_matrix(fit_pbmc)
plot_consensus_matrix(fit_pbmc, sweep_value = 1.5)
```

`sweep_value` pins a resolution in the consensus, violin and scatter
plots and in `get_labels()`. On
the UMAP, at the selected resolution and at the finest one:

```{r pbmc-scatter}
umap <- SeuratObject::Embeddings(pbmc, reduction = "umap")
plot_cluster_scatter(fit_pbmc, embedding = umap)
plot_cluster_scatter(fit_pbmc, embedding = umap, sweep_value = 1.5)
plot_cluster_violin(fit_pbmc, sweep_value = 0.25)
plot_cluster_violin(fit_pbmc, sweep_value = 1.5)
```

The results table gains the sweep columns:

```{r pbmc-results}
results_pbmc <- estimator_results(fit_pbmc)
results_pbmc[, c("method_label", "resolution", "sweep_rank", "n_clusters_observed",
                 "ari_stability", "ari_generalizability")]
```

`sweep_rank` orders the values from coarse to fine, and the selection
rules work on it, so they behave the same on every axis.
`n_clusters_observed` is the mean number of clusters over the resamples.

## Labels and Seurat's clusters

```{r pbmc-labels}
labels_pbmc <- get_labels(fit_pbmc)
table(CARVE = labels_pbmc, Seurat = pbmc$seurat_clusters)
mclust::adjustedRandIndex(labels_pbmc, pbmc$seurat_clusters)
table(get_labels(fit_pbmc, consensus_k = 4))
```

`consensus_k` cuts the selected consensus matrix into another number of
clusters, to see how the cells group at a coarser level.

## Storing the results in the object

`attach_results()` writes the selected configuration into the Seurat
object: the labels and the per-cell scores into the cell metadata, and the
parameters, the results table and the consensus matrix into
`pbmc@misc$carve`.

```{r pbmc-attach}
pbmc <- attach_results(pbmc, fit_pbmc)
head(pbmc@meta.data[, c("seurat_clusters", "carve", "carve_stability", "carve_generalizability")])
Seurat::DimPlot(pbmc, group.by = "carve")
plot_diagnostic_scatter(pbmc, basis = "umap")
```

`run_carve()` does both steps in one call. On a SingleCellExperiment made
from the same object, with the same seed, it gives the same labels:

```{r pbmc-sce}
sce <- Seurat::as.SingleCellExperiment(pbmc)
SingleCellExperiment::reducedDimNames(sce)
sce <- run_carve(sce, resolution = resolutions, n_dims = 10, n_jobs = n_jobs)
identical(as.character(sce$carve), as.character(pbmc$carve))
plot_cluster_scatter(sce, basis = "UMAP")
```

## Ambiguous cells

```{r pbmc-noise}
noisy <- get_labels(fit_pbmc, noise_labels = TRUE, noise_quantile = 0.05)
table(noisy)
pbmc$carve_noise <- factor(noisy)
Seurat::DimPlot(pbmc, group.by = "carve_noise")
```

`noise_labels = TRUE` labels -1 the cells whose stability is in the
lowest 5 percent and more than 0.05 below the median.

## HDBSCAN

HDBSCAN takes a minimum cluster size and labels cells in sparse regions
-1. It needs the dbscan package.

```{r pbmc-hdbscan}
pcs <- SeuratObject::Embeddings(pbmc, reduction = "pca")[, 1:10]
hdbscan_grid <- list(estimator_grid(HDBSCAN, min_cluster_size = c(10, 20, 40, 80)))
fit_hdbscan <- carve(pcs, estimator_param_grids = hdbscan_grid, n_jobs = n_jobs, random_state = 0)
sweep_spec(fit_hdbscan)
estimator_results(fit_hdbscan)[, c("min_cluster_size", "n_clusters_observed", "noise_fraction", "ari_stability")]
get_sweep_value(fit_hdbscan)
```

From the grid, CARVE infers that `min_cluster_size` is swept and that
larger values give fewer clusters. `noise_fraction` is the share of cells
labeled noise, averaged over the resamples. By default, `noise_policy =
"drop"` leaves noise cells out of each resample's scores.

## Graph methods on their own, and neighborhood sizes

`LeidenClustering()` and `LouvainClustering()` are plain functions:

```{r pbmc-graph}
table(LeidenClustering(pcs, resolution = 0.5, random_state = 0))
table(LouvainClustering(pcs, resolution = 0.5, random_state = 0))
```

A grid can vary the graph as well as the resolution. CARVE infers the
resolution sweep from the grid:

```{r pbmc-neighbors}
neighbor_grid <- list(estimator_grid(LeidenClustering, resolution = resolutions, n_neighbors = c(10, 15, 30)))
fit_neighbors <- carve(pcs, estimator_param_grids = neighbor_grid, n_jobs = n_jobs, random_state = 0)
sweep_spec(fit_neighbors)
plot_metric_over_n_clusters(fit_neighbors)
```

## Randomized preprocessing

The principal components above are one choice of preprocessing. With
`randomize_preprocessing = TRUE`, CARVE draws a pipeline for each resample
and fits it on the subsamples, here on the log-normalized expression of
the 2,000 variable genes. The default backend forks the R session. After
Seurat's PCA had run in this session, the forked workers crashed when they
computed a PCA, so this fit passes a socket backend through `BPPARAM`:

```{r pbmc-randomized}
genes <- SeuratObject::VariableFeatures(pbmc)
expression <- t(as.matrix(SeuratObject::LayerData(pbmc, layer = "data")[genes, ]))
dim(expression)
fit_random_pbmc <- carve(
  expression, randomize_preprocessing = TRUE, n_resamples = 60, n_jobs = n_jobs, random_state = 0,
  BPPARAM = BiocParallel::SnowParam(workers = n_jobs),
  estimator_param_grids = list(estimator_grid(LeidenClustering, resolution = c(0.25, 0.5, 1))),
  normalization_options = list(preprocessing_option(Identity), preprocessing_option(StandardScaler)),
  dim_reduction_options = list(preprocessing_option(PCA, n_components = c(10, 20, 30)))
)
plot_metric_by_pipeline(fit_random_pbmc)
```

`pipeline_from_spec()` applies a pipeline to all the cells, for example
the one with the highest stability:

```{r pbmc-pipeline}
pipeline_results <- preprocessing_results(fit_random_pbmc)
best <- pipeline_results$pipeline[which.max(pipeline_results$ari_stability)]
best
embedding <- pipeline_from_spec(preprocessing_pipelines(fit_random_pbmc)[[best]], random_state = 0)(expression)
dim(embedding)
```

# Session information

This rendering took
`r round(as.numeric(difftime(Sys.time(), tutorial_start, units = "mins")))`
minutes on `r n_jobs` cores.

```{r session}
sessionInfo()
```
