---
title: "CARVE tutorial for R"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
tutorial_start <- Sys.time()
```

This tutorial goes through the CARVE R package, on simulated data in
part 1 and on the PBMC 3k single-cell data set in part 2. It follows the
Python notebooks `Tutorial.ipynb` and `Resolution_Tutorial.ipynb` in this
folder and fits at full size, with the default 100 resamples. The
package's vignettes explain the same steps on smaller data:
`vignette("CARVE", package = "CARVE")` for the scores and the selection
rules, `vignette("single-cell", package = "CARVE")`,
`vignette("customizing", package = "CARVE")` and
`vignette("python-users", package = "CARVE")`.

Part 2 needs the Seurat, SeuratObject, dbscan and mclust packages, and
downloads 7 MB from 10x Genomics into `data/` on its first run.

```{r packages}
library(CARVE)

n_jobs <- max(1L, parallel::detectCores() - 1L)
n_jobs
```

`n_jobs` is the number of cores each fit may use.

# Part 1: simulated data

## Quick start

Two pairs of interleaved half-moons, 1,000 points in four groups. The
groups are curved, which k-means handles poorly.

```{r moons, fig.width = 6, fig.height = 4}
make_moons <- function(n, noise) {
  n_outer <- n %/% 2
  n_inner <- n - n_outer
  t_outer <- seq(0, pi, length.out = n_outer)
  t_inner <- seq(0, pi, length.out = n_inner)
  X <- rbind(cbind(cos(t_outer), sin(t_outer)),
             cbind(1 - cos(t_inner), 1 - sin(t_inner) - 0.5))
  X + matrix(rnorm(2 * n, sd = noise), ncol = 2)
}

set.seed(42)
moons <- rbind(make_moons(500, noise = 0.12),
               make_moons(500, noise = 0.10) + matrix(c(2, 1.5), 500, 2, byrow = TRUE))
moons_truth <- rep(1:4, each = 250)
plot(moons, col = moons_truth, pch = 19, cex = 0.5, asp = 1, xlab = "", ylab = "")
```

```{r moons-fit}
fit_moons <- carve(moons, n_clusters = 2:8, n_jobs = n_jobs, random_state = 42)
fit_moons
get_k(fit_moons)
moons_labels <- get_labels(fit_moons)
table(moons_labels, moons_truth)
moons_ari <- mclust::adjustedRandIndex(moons_labels, moons_truth)
moons_ari
```

CARVE selects `r get_k(fit_moons)` clusters, and the labels have an ARI of
`r round(moons_ari, 2)` with the four true groups. Each of the two clusters
holds one pair of moons. The pairs are far apart and resampling finds that
split stable, while the moons inside a pair are not recovered as separate
clusters by any of the three default methods. `get_labels(k = 4)` splits
each pair in two without following the moons, and its ARI with the truth is
`r round(mclust::adjustedRandIndex(get_labels(fit_moons, k = 4), moons_truth), 2)`.

```{r moons-plots}
plot_metric_over_n_clusters(fit_moons)
plot_cluster_scatter(fit_moons)
```

## How CARVE scores a configuration

A configuration is a clustering method, its parameters and a value of the
swept parameter, here the number of clusters. For each of `n_resamples`
resamples, CARVE draws two overlapping subsamples of `subsample_ratio` of
the samples (0.618 by default) and clusters both. Stability is the
adjusted Rand index (ARI) between the two clusterings on the samples they
share. For generalizability, a random forest trained on one subsample's
clusters predicts the clusters of the samples held out of it, and the ARI
compares the prediction with the clusters those samples were given. After
the resamples, CARVE builds each configuration's consensus matrix, the
share of the subsamples drawing two samples in which they fell in the
same cluster, and computes per-sample scores from it.

```{r results}
results <- estimator_results(fit_moons)
names(results)
results[results$n_clusters == 4,
        c("method_label", "ari_stability", "ari_generalizability",
          "consensus_pac_stability", "consensus_gini_stability", "accuracy_generalizability")]
```

`ari_stability` and `ari_generalizability` are means over the resamples,
with standard errors (`_se`) and 5th and 95th percentiles (`_lower`,
`_upper`). `consensus_pac_stability` is one minus the proportion of
ambiguous clustering, the share of consensus entries between 0.05 and 0.95,
so higher is better, as for every score in the table.
`consensus_gini_stability` and `consensus_ce_stability` average the
per-sample stability scores, and `accuracy_generalizability` the
per-sample accuracy of the classifier on held-out samples.

## Plots

The eight plots return ggplot objects. The metric plot draws one line per
method over the swept values, and the dashed line marks the selection:

```{r plot-metric}
plot_metric_over_n_clusters(fit_moons, measure = "generalizability")
```

`plot_n_clusters_over_sweep()` draws the number of clusters each method
returned, averaged over the resamples. It needs a sweep over a parameter
other than `n_clusters`, such as a resolution, where the count is an
outcome; on an `n_clusters` sweep the count is the swept value itself and
the function stops with an error saying so. Part 2 uses it.

The consensus heatmap orders the samples by cluster. `k` shows another
number of clusters:

```{r plot-consensus, fig.width = 5.5, fig.height = 5}
plot_consensus_matrix(fit_moons)
plot_consensus_matrix(fit_moons, k = 6)
```

The per-sample plots show the stability of each sample (`source = "gini"`
or `"ce"`) or the share of the draws holding it out in which the
classifier predicted its cluster (`source = "accuracy"`):

```{r plot-per-sample}
plot_cluster_boxplot(fit_moons)
plot_cluster_violin(fit_moons, source = "accuracy")
plot_diagnostic_scatter(fit_moons)
```

`plot_metric_by_pipeline()`, the eighth, is in the section on randomized
preprocessing below.

## Selecting a configuration and labeling the samples

```{r selecting}
get_k(fit_moons, measure = "generalizability")
get_sweep_value(fit_moons)
moons_estimator <- get_estimator(fit_moons)
attributes(moons_estimator)[c("estimator", "params")]
table(get_labels(fit_moons, k = 4), moons_truth)
```

`get_estimator()` returns the selected method as a function of the data
and a seed, and `get_labels(k = 4)` cuts the best configuration at four
clusters instead of the selected one. Labels run from 1 to the number of
clusters.

## Easy and hard data

Three well-separated groups, and five that overlap:

```{r easy-hard}
make_blobs <- function(n_per, centers, sd) {
  truth <- rep(seq_len(nrow(centers)), each = n_per)
  noise <- matrix(rnorm(length(truth) * ncol(centers), sd = sd), ncol = ncol(centers))
  list(X = centers[truth, ] + noise, truth = truth)
}

set.seed(0)
easy <- make_blobs(100, rbind(c(0, 0), c(8, 0), c(4, 7)), sd = 1)
hard <- make_blobs(100, rbind(c(0, 0), c(3, 0), c(1.5, 2.6), c(4.5, 2.6), c(3, 5.2)), sd = 1.2)
fit_easy <- carve(easy$X, n_clusters = 2:8, n_jobs = n_jobs, random_state = 0)
fit_hard <- carve(hard$X, n_clusters = 2:8, n_jobs = n_jobs, random_state = 0)
c(easy = get_k(fit_easy), hard = get_k(fit_hard))
```

```{r easy-hard-plots}
plot_metric_over_n_clusters(fit_easy) + ggplot2::labs(title = "Three separated groups")
plot_metric_over_n_clusters(fit_hard) + ggplot2::labs(title = "Five overlapping groups")
```

On the easy data the scores of all three methods peak at the true three
groups, at an ARI near 1. On the hard data the scores are lower and the
methods disagree, and the best score is at two clusters, not at the five
true groups: groups that overlap this much leave no five-cluster
structure for the resamples to reproduce.

## The selection rules

The same hard fit under each measure and rule:

```{r rules}
measures <- c("stability", "generalizability", "average")
rules <- c("max", "1se", "quantile")
sapply(rules, function(rule) {
  sapply(measures, function(measure) get_k(fit_hard, measure = measure, rule = rule))
})
```

`"max"` takes the highest mean score. `"1se"` takes the largest number of
clusters whose score is within one standard error of the highest, and
`"quantile"` the largest within the 5th to 95th percentiles of the best
configuration's resample scores, so both lean toward finer clusterings
when scores are close. On this fit `"max"` and `"1se"` give two clusters
under every measure. `"quantile"` gives
`r get_k(fit_hard, measure = "stability", rule = "quantile")` for stability
and `r get_k(fit_hard, measure = "generalizability", rule = "quantile")`
for generalizability, so the two scores can point to different numbers of
clusters when the rule is looser.

## Two clusters

Four groups in a row can be split into two halves the same way in every
resample, so two clusters can score as well as four:

```{r not-two}
set.seed(1)
line <- make_blobs(75, cbind(c(0, 3, 6, 9), 0), sd = 0.7)
fit_line <- carve(line$X, n_clusters = 2:8, n_jobs = n_jobs, random_state = 1)
c(max = get_k(fit_line, rule = "max"),
  max_not_two = get_k(fit_line, rule = "max", not_two = TRUE),
  one_se = get_k(fit_line))
```

```{r not-two-plots, fig.width = 5.5, fig.height = 5}
plot_consensus_matrix(fit_line, k = 2)
plot_consensus_matrix(fit_line, rule = "max", not_two = TRUE)
```

The two-cluster consensus matrix looks clean, but it says little about
the four groups. `not_two = TRUE` leaves two clusters out of the
selection.

## Custom grids

`estimator_grid()` sets the methods and the values of their parameters:

```{r grids}
grids <- list(
  estimator_grid(KMeans, n_clusters = 2:8, n_init = 10),
  estimator_grid(AgglomerativeClustering, n_clusters = 2:8, linkage = c("ward", "single")),
  estimator_grid(SpectralClustering, n_clusters = 2:8, n_neighbors = c(5, 15))
)
fit_grid <- carve(moons, estimator_param_grids = grids, n_jobs = n_jobs, random_state = 42)
unique(estimator_results(fit_grid)$method_label)
get_k(fit_grid)
plot_metric_over_n_clusters(fit_grid)
```

## Randomized preprocessing

k-means does not follow the curved shapes of the moons in their own
coordinates. With
`randomize_preprocessing = TRUE`, each resample draws a normalization and
a dimensionality reduction and clusters the transformed subsamples; the
classifier still trains on the raw coordinates.

```{r randomized}
fit_random <- carve(
  moons, randomize_preprocessing = TRUE, n_resamples = 60, n_jobs = n_jobs, random_state = 42,
  estimator_param_grids = list(estimator_grid(KMeans, n_clusters = 2:8)),
  normalization_options = list(preprocessing_option(Identity), preprocessing_option(StandardScaler)),
  dim_reduction_options = list(preprocessing_option(Identity), preprocessing_option(TSNE, perplexity = c(30, 50)))
)
plot_metric_by_pipeline(fit_random)
pipeline_results <- preprocessing_results(fit_random)
pipeline_results[pipeline_results$n_clusters == 4,
                 c("pipeline", "n_resamples", "ari_stability", "ari_generalizability")]
```

`preprocessing_results()` splits each configuration's scores by pipeline.
This run uses 60 resamples, because every resample runs t-SNE on its
subsamples, so each pipeline's scores rest on 5 to 15 resamples
(`n_resamples` in the table) and are noisy. At four clusters the pipelines
with t-SNE do not score higher than the identity pipelines on these data.

## Saving a fit

```{r save}
path <- file.path(tempdir(), "fit_moons.rds")
saveRDS(fit_moons, path)
reloaded <- readRDS(path)
identical(get_labels(reloaded), moons_labels)
round(file.size(path) / 2^20, 1)
```

The size is in MB. A fit keeps the data, so the scatter plots work on a
fit read back from disk; Python's `save()` leaves the data out unless
`include_data=True`.

## Reference labels

Cluster numbers are arbitrary. `reference_labels` renames the clusters to
match another labeling with the same number of clusters, here the truth:

```{r reference}
table(get_labels(fit_moons, k = 4, reference_labels = moons_truth), moons_truth)
```

`carve(reference_labels = )` sets the labeling for every `get_labels()`
call on the fit.

## Parallel runs

A run clusters every configuration on every resample, so its cost grows
with the number of configurations times `n_resamples`. `n_jobs` spreads
each configuration's resamples over worker processes; the results do not
depend on it.

```{r timing}
blob_grid <- list(estimator_grid(KMeans, n_clusters = 2:8))
time_serial <- system.time(fit_serial <- carve(hard$X, estimator_param_grids = blob_grid, random_state = 0))
time_workers <- system.time(fit_workers <- carve(hard$X, estimator_param_grids = blob_grid, n_jobs = n_jobs, random_state = 0))
c(serial = time_serial[["elapsed"]], workers = time_workers[["elapsed"]])
all.equal(estimator_results(fit_serial), estimator_results(fit_workers))
```

Starting the workers has a cost of its own, so the gain grows with the
work in each resample.

# Session information

This rendering took
`r round(as.numeric(difftime(Sys.time(), tutorial_start, units = "mins")))`
minutes on `r n_jobs` cores.

```{r session}
sessionInfo()
```
